Exploring how changes in cause of death have narrowed the

### 2017-02-14 Manuscript/Manuscript_COD_decomp.Rmd
more changes
first set of JK revisions

---
title: Exploring how changes in cause of death have narrowed the
  black-white life expectancy gap over forty-five years in the United States.
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

Kathryn T Morrison, MSc

@@

The difference in life expectancy between black and whites in the US has decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.

**Methods and Findings**
Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

@@

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race. Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data was derived by the NCHS using intercensal bridged-race population estimates. This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata. However, death counts of exactly 0 or greater than 9 are known precisely.

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the blacks in the United States in 2010 [ref]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].

@@

The data produced two statistical issues that were addressed by our statistical methods. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.

To decompose the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.

@@

For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectnacy between blacks and whites according to cause of death. To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era. An example of the calculation is found in the eAppendix.

*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each strata. We performed all calculated on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th qunatiles as the lower and upper bound for the 95% CI. Anything about thinning?

All data analyses were performed using R version 3.3.1. The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.

@@

**Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines. This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap. The length of these coloured bars represent the number of years attributable to the cause.

@@

In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `. Injuries was the largest contributor towards narrowing the gap during this time period, as both races underwent decreased rates of injury-related mortality, but blacks experienced larger decreases to their rates, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**). Minnesota and Nebraska are exceptions, where both states had relatively low rates of injuries among blacks in 1969 and smaller gaps. All other causes also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries. In New York, all other causes made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**). Communicable diseases modestly contributed to narrowing the gap, except in Arizona and Minnesota. Generally, cardivascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites. Cancer generally contributed towards increasing the gap, and largely reflects increased cancer-related mortality for both races, with blacks undergoing steeper changes. In most states, non-comunicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.

For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r max.cause5`).

The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states. Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)`, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)`, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)`, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)`, `r abs(round(min4,1))` due to `r min.cause4`). For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widens the gap), so these gap increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.

During the second time period, from 1983 to 1993, the life expectancy gap for males in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the analysed `r dim(bw_dat.males)[1]` widened (**Fig 2**). For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened risk of dying from HIV/AIDS in black males during the height of the AIDS epidemic. While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher. Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana. For many states, injury-related mortality increased for blacks but not for whites during this time period. Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps. Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions. All other causes contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL]. Most notably, the Disrict of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded). By 2013, it ranged from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` in `r sum4[dim(sum4)[1]-1, ]$state` when DC is excluded). During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardivascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to cardiovascular disease. As you can see, in some states CVD contributed to excacerbating the gap while in most others it makes a large contribution to narrowing the gap.

>>>>

---
title: Exploring how changes in cause of death have narrowed the
  black-white life expectancy gap in the United States.
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

Kathryn T Morrison, PhD

@@

The overall difference in life expectancy between black and whites in the US has generally decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York experiencing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in narrower life expectancy gaps in some states and  wider gaps in other states.

**Methods and Findings**
Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A Bayesian autoregressive model was used to model smoothed mortality rates over time within each stratum, and a truncated Poisson model was used to impute death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each stratum using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

@@

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race. Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined. Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].

@@

The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). These models were run separately for every combination of state, race, and sex, and produced death counts that were smoothed over time for every cause of death within each age group.

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.

@@

For each era, we calculated the degree that the life expectancy gap has narrowed or widened and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death. To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era. An example of the calculation is found in the eAppendix.

*To re-write*: To calculate 95% credible intervals, we sampled 1000 times from the posterior distribution to create 1000 samples for each stratum. We performed all calculations on each sample and took the median of the estimates as the estimate and the 2.5 and 97.5th quantiles as the lower and upper bound for the 95% CI. Anything about thinning?

All data analyses were performed using R version 3.3.1. The Bayesian smoothing was performed using `JAGS` (version X.X), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.

@@

**Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines along the x-axis. This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap. The length of these coloured bars represent the number of years attributable to the cause along the x-axis.

@@

In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `. Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**). A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps. The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries. In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**). Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota. Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites. Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes. In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.

For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` of the `r dim(bw_dat.males)[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` years, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)` years, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)` years, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)` years, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)` years, `r round(max5,1)` due to `r max.cause5`).

The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states. Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)` years, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)` years, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)` years, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)` years, `r abs(round(min4,1))` due to `r min.cause4`). For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widened the gap), so these disparity increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.

During the second time period, from 1983 to 1993, the life expectancy gap for males widened in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the `r dim(bw_dat.males)[1]` states (**Fig 2**). For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened burden from HIV/AIDS in black males during the height of the AIDS epidemic. While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher. Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana. For many states, injury-related mortality increased for blacks but not for whites during this time period. Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps. Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions. The residual category contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL]. Most notably, the District of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase in the gap of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

- However, increases were found in Arizona, Oregon, Minnesota, Washington, and Wisconsin, largely due to CVD. As you can see, in some states CVD contributed to excacerbating the gap while in most others it makes a large contribution to narrowing the gap.
